---
title: 'The Road to 270: Democratic Campaigning Strategy for the 2024 US General Election'
date: "April 23, 2024"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: none
    self_contained: false
    css: styles.css
---

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
# Function to check and install missing packages
install_if_missing <- function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
    library(package, character.only = TRUE)
  }
}

# List of required packages
packages <- c(
  "countrycode", "dplyr", "DT", "ggplot2", "gridExtra", "leaflet", 
  "maps", "patchwork", "readr", "rworldmap", "showtext", 
  "sp", "tidyverse"
)

# Ensure all packages are installed and loaded
lapply(packages, install_if_missing)

# Load libraries explicitly
library(countrycode)
library(dplyr)
library(DT)
library(ggplot2)
library(gridExtra)
library(leaflet)
library(maps)
library(patchwork)
library(readr)
library(rworldmap)
library(showtext)
library(sp)
library(tidyverse)

# Set the file path
url <- "https://raw.githubusercontent.com/annab0503/CEIR/refs/heads/main/Writing%20Sample/Original%20Dataframes/2016-2022%20Validated%20Voter%20Detailed%20Tables%20-%20Sheet%201.csv"

# Read the CSV file from the URL
data <- read.csv(url, skip = 3)

# Enable the showtext package
showtext_auto()

# Add Google Fonts
font_add_google("Bodoni Moda", "bodoni_moda")  # Alias: bodoni_moda
font_add_google("Montserrat", "montserrat")     # Alias: montserrat
```

# INTRODUCTION
## *The Democratic Presidential campaign **must** establish their strategy to secure electoral victory.*
Previous research shows that voter turnout shapes election outcomes. When voter turnout is high, it often indicates a high level of engagement and enthusiasm among the electorate, which can lead to more representative results. Conversely, low voter turnout can skew outcomes, as it may reflect apathy, disenchantment, or barriers to participation among certain demographics. Elections with low turnout rates risk being decided by a smaller, potentially less diverse subset of the population, which can distort the democratic process. Furthermore, variations in voter turnout across different regions or demographic groups can significantly impact the final results, influencing everything from the composition of legislative bodies to the outcome of specific ballot measures or initiatives. Therefore, understanding and addressing factors that affect voter turnout is essential for ensuring fair and representative democratic elections.

By analyzing voting patterns from the previous four National election cycles (2016-2022), the Democratic Presidential campaign can obtain valuable insight on key voter demographics, including voter turnout and preferences, which can help information strategic decisions regarding messaging, campaign outreach, and policy priorities to effectively engage and mobilize supporters in the upcoming election.

Following the primary elections earlier this year, we are now certain that the 2024 Presidential Election will be a Biden-Trump rematch. While the political environment has changed since their previous face-off in 2020, we can gain valuable insight on general trends in among voter demographics. In a time of immense polarization where both candidates are are well-established incumbents, the Democratic campaign can not rely on vote flipping. *Therefore, with the U.S. General Election only about six months year away, **Democrats need to consolidate a strong voting base that will turn out to on election day if they want to ensure a win.***

# METHODOLOGY
Following each election, the Pew Research Center surveys U.S. adults online to study trends in voter turnout and choices. Using this survey data, I am analyzing subsets of voters with low voter turnout for general trends that may be expected to continue in 2024.

Variables in my (cleaned) dataset include:

| Variable             | Description                                                    |
|----------------------|----------------------------------------------------------------|
| Type of Voter        | Categorical groups of surveyed voters, by key characteristics  |
| Year                 | The year of the election                                       |
| Type of Election     | The type of the election, either House of Reps or Presidential |
| D Vote               | Percentage of vote received by Democratic Party                |
| D Voter Share        | Percentage of all Democratic Party votes from Type of Voter    |
| D-R Margin           | Difference between Democratic and Republican Party votes       |
| Total Nonvoter Share | Percentage of all nonvoters from Type of Voter                 |
| R Vote               | Percentage of vote received by Republican Party                |
| R Voter Share        | Percentage of all Republican Party votes from Type of Voter    |
| Total Nonvoter Share | Percentage of all voters from Type of Voter                    |

The Pew Research Center's most recent report using this data can be found at [www.pewresearch.org](https://www.pewresearch.org/politics/2023/07/12/voter-turnout-2018-2022/){.btn .btn-primary}

I originally accessed the survey data on March 21, 2024.

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
data
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
# General Cleaning

data <- data %>% #Use dplyr piping to clean data
  # Rename columns
  rename(
    "Type of Voter" = X,
    "2022 D House Vote" = X2022.Democratic.House.vote..official.result.48..,
    "2022 R House Vote" = X2022.Republican.House.vote..official.result.51..,
    "2022 D-R House Margin" = X2022.Margin..Dem.Rep.,
    "2020 D Pres Candidate Vote" = X2020.Biden.vote..official.result.51..,
    "2020 R Pres Candidate Vote" = X2020.Trump.vote..official.result.47..,
    "2020 D-R Pres Candidate Margin" = X2020.Margin..Biden.Trump.,
    "2018 D House Vote" = X2018.Democratic.House.vote..official.result.53..,
    "2018 R House Vote" = X2018.Republican.House.vote..official.result.44..,
    "2018 D-R House Margin" = X2018.Margin..Dem.Rep.,
    "2016 D Pres Candidate Vote" = X2016.Clinton.vote..official.result.48..,
    "2016 R Pres Candidate Vote" = X2016.Trump.vote..official.result.46..,
    "2016 D-R Pres Candidate Margin" = X2016.Margin..Clinton.Trump.,
    "X14" = X2022.Vote.choice.margin.of.error......at.95..CI.,
    "X15" = X2020.Vote.choice.margin.of.error......at.95..CI.,
    "X16" = X2018.Vote.choice.margin.of.error......at.95..CI.,
    "X17" = X2016.Vote.choice.margin.of.error......at.95..CI.,
    "X18" = Group.as.a.share.of.all.2022.voters,
    "X19" = Group.as.a.share.of.all.2020.voters,
    "X20" = Group.as.a.share.of.all.2018.voters,
    "X21" = Group.as.a.share.of.all.2016.voters,
    "2022 R Voter Share" = Group.as.a.share.of.2022.Republican.voters,
    "2022 D Voter Share" = Group.as.a.share.of.2022.Democratic.voters,
    "2020 R Pres Candidate Voter Share" = Group.as.a.share.of.2020.Trump.voters,
    "2020 D Pres Candidate Voter Share" = Group.as.a.share.of.2020.Biden.voters,
    "2018 R Voter Share" = Group.as.a.share.of.2018.Republican.voters,
    "2018 D Voter Share" = Group.as.a.share.of.2018.Democratic.voters,
    "2016 R Pres Candidate Voter Share" = Group.as.a.share.of.2016.Trump.voters,
    "2016 D Pres Candidate Voter Share" = Group.as.a.share.of.2016.Clinton.voters,
    "2022 Total Voter Share" = Group.as.a.share.of.all.2022.voters.1,
    "2022 Total Nonvoter Share" = Group.as.a.share.of.all.2022.nonvoters,
    "2020 Total Voter Share" = Group.as.a.share.of.all.2020.voters.1,
    "2020 Total Nonvoter Share" = Group.as.a.share.of.all.2020.nonvoters,
    "2018 Total Voter Share" = Group.as.a.share.of.all.2018.voters.1,  
    "2018 Total Nonvoter Share" = Group.as.a.share.of.all.2018.nonvoters,  
    "2016 Total Voter Share" = Group.as.a.share.of.all.2016.voters.1,  
    "2016 Total Nonvoter Share" = Group.as.a.share.of.all.2016.nonvoters,  
    "X38" = X.1,  
    "X39" = rep_18_22,  
    "X40" = dem_18_22
  ) %>%
  # Remove unnecessary/duplicate columns
  select(-c(X14:X21, X38:X40)) %>%
  # Replace empty strings with NA
  mutate(across(everything(), ~ifelse(. == "", NA, .))) %>%
  # Replace dashes with NA
  mutate(across(everything(), ~ifelse(. == "-", NA, .))) %>%
  # Convert all column values to integers
   mutate_at(vars(2:ncol(.)), as.integer) %>%
 # Reorder columns alphabetically
  select(sort(names(.))) %>%
  select(all_of("Type of Voter"), everything())

# Split original dataframe into its individual subsets:

# Find indices of rows with all NA or empty values
empty_row_indices <- which(apply(data, 1, function(x) all(is.na(x) | x == "")))

# Initialize a list to store subsets of data
subsets <- list()

# Loop through each row in the data
for (i in 1:nrow(data)) {
  # Check if the current row is in the empty row indices
  if (i %in% empty_row_indices) {
    # If it's empty, store the next row's "Type of Voter" value as the current name
    current_name <- as.character(data[i + 1, "Type of Voter"])
  } else {
    # If the row is not empty
    if (!is.na(current_name)) {
      # Extract the current row as a subset of data
      subset_data <- data[i, , drop = FALSE]
      # Check if the current name exists as a key in the subsets list
      if (!(current_name %in% names(subsets))) {
        # If not, add the subset data as a new list element
        subsets[[current_name]] <- subset_data
      } else {
        # If yes, append the subset data to the existing list element
        subsets[[current_name]] <- rbind(subsets[[current_name]], subset_data)
      }
    }
  }
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
# VISUALIZATION 1 CLEANING

gender_data <- rbind(subsets[["Gender"]], subsets[["Gender and race"]], subsets[["Gender and Marital"]], subsets[["Gender and education"]], subsets[["Gender and age"]]) 

gender_data <- gender_data %>%
  mutate(`Type of Voter` = ifelse(grepl("Gender", `Type of Voter`), NA, `Type of Voter`)) %>%
  filter_all(any_vars(!is.na(.)))

# Split the dataframe by election/year for individual cleaning
data1 <- gender_data[, c("Type of Voter", "2016 D Pres Candidate Vote", "2016 D Pres Candidate Voter Share", 
                  "2016 D-R Pres Candidate Margin", "2016 Total Nonvoter Share", 
                  "2016 R Pres Candidate Vote", "2016 R Pres Candidate Voter Share", 
                  "2016 Total Voter Share")]
data2 <- gender_data[, c("Type of Voter", "2018 D House Vote", "2018 D Voter Share", 
                  "2018 D-R House Margin", "2018 Total Nonvoter Share", 
                  "2018 R House Vote", "2018 R Voter Share", 
                  "2018 Total Voter Share")]

data3 <- gender_data[, c("Type of Voter", "2020 D Pres Candidate Vote", "2020 D Pres Candidate Voter Share", 
                  "2020 D-R Pres Candidate Margin", "2020 Total Nonvoter Share", 
                  "2020 R Pres Candidate Vote", "2020 R Pres Candidate Voter Share", 
                  "2020 Total Voter Share")]

data4 <- gender_data[, c("Type of Voter", "2022 D House Vote", "2022 D Voter Share", 
                  "2022 D-R House Margin", "2022 Total Nonvoter Share", 
                  "2022 R House Vote", "2022 R Voter Share", 
                  "2022 Total Voter Share")]

# Add a "Type of Election" column
data1 <- cbind(data1[, 1, drop = FALSE], 'Type of Election' = 'Presidential', data1[, -1])
data2 <- cbind(data2[, 1, drop = FALSE], 'Type of Election' = 'House', data2[, -1])
data3 <- cbind(data3[, 1, drop = FALSE], 'Type of Election' = 'Presidential', data3[, -1])
data4 <- cbind(data4[, 1, drop = FALSE], 'Type of Election' = 'House', data4[, -1])

# Remove type of election from all cells in the title row
colnames(data1) <- gsub("Pres Candidate ", "", colnames(data1))
colnames(data2) <- gsub("House ", "", colnames(data2))
colnames(data3) <- gsub("Pres Candidate ", "", colnames(data3))
colnames(data4) <- gsub("House ", "", colnames(data4))

# Add a "Year" column
data1 <- cbind(data1[, 1, drop = FALSE], Year = 2016, data1[, -1])
data2 <- cbind(data2[, 1, drop = FALSE], Year = 2018, data2[, -1])
data3 <- cbind(data3[, 1, drop = FALSE], Year = 2020, data3[, -1])
data4 <- cbind(data4[, 1, drop = FALSE], Year = 2022, data4[, -1])

# Remove year from all cells in the title row
colnames(data1) <- gsub("2016 ", "", colnames(data1))
colnames(data2) <- gsub("2018 ", "", colnames(data2))
colnames(data3) <- gsub("2020 ", "", colnames(data3))
colnames(data4) <- gsub("2022 ", "", colnames(data4))

# Merge cleaned dataframes
clean_gender_data <- rbind(data1, data2, data3, data4)

# Subset the dataframe by gender from the "Type of Voter" column
male_voters <- clean_gender_data[grepl("\\b(Men| men)\\b", clean_gender_data$`Type of Voter`, ignore.case = TRUE), ]
female_voters <- clean_gender_data[grepl("Women", clean_gender_data$`Type of Voter`, ignore.case = TRUE), ]

# Add a "Gender" column "Gender" to subset dataframes
male_voters$Gender <- "Male"
female_voters$Gender <- "Female"

# Remove gender indicator from "Type of Voter" column in the subset dataframes
male_voters$`Type of Voter` <- gsub("Men, ", "", male_voters$`Type of Voter`, ignore.case = TRUE)
female_voters$`Type of Voter` <- gsub("Women, ", "", female_voters$`Type of Voter`, ignore.case = TRUE)

# Replace remaining cells with "Total"
male_voters$`Type of Voter`[male_voters$`Type of Voter` == "Men"] <- "Total"
female_voters$`Type of Voter`[female_voters$`Type of Voter` == "Women"] <- "Total"

gender_data_final <- rbind(male_voters, female_voters)
gender_data_final <- gender_data_final %>%
  select('Gender', everything())
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
# VISUALIZATION 2 CLEANING

age_data <- subsets[["Age (six-way)"]]
age_data <- age_data %>%
  mutate(`Type of Voter` = ifelse(grepl("Age", `Type of Voter`), NA, `Type of Voter`)) %>%
  filter_all(any_vars(!is.na(.)))

# Split the dataframe by election/year for individual cleaning
data1 <- age_data[, c("Type of Voter", "2016 D Pres Candidate Vote", "2016 D Pres Candidate Voter Share", 
                  "2016 D-R Pres Candidate Margin", "2016 Total Nonvoter Share", 
                  "2016 R Pres Candidate Vote", "2016 R Pres Candidate Voter Share", 
                  "2016 Total Voter Share")]
data2 <- age_data[, c("Type of Voter", "2018 D House Vote", "2018 D Voter Share", 
                  "2018 D-R House Margin", "2018 Total Nonvoter Share", 
                  "2018 R House Vote", "2018 R Voter Share", 
                  "2018 Total Voter Share")]

data3 <- age_data[, c("Type of Voter", "2020 D Pres Candidate Vote", "2020 D Pres Candidate Voter Share", 
                  "2020 D-R Pres Candidate Margin", "2020 Total Nonvoter Share", 
                  "2020 R Pres Candidate Vote", "2020 R Pres Candidate Voter Share", 
                  "2020 Total Voter Share")]

data4 <- age_data[, c("Type of Voter", "2022 D House Vote", "2022 D Voter Share", 
                  "2022 D-R House Margin", "2022 Total Nonvoter Share", 
                  "2022 R House Vote", "2022 R Voter Share", 
                  "2022 Total Voter Share")]

# Add a "Type of Election" column
data1 <- cbind(data1[, 1, drop = FALSE], 'Type of Election' = 'Presidential', data1[, -1])
data2 <- cbind(data2[, 1, drop = FALSE], 'Type of Election' = 'House', data2[, -1])
data3 <- cbind(data3[, 1, drop = FALSE], 'Type of Election' = 'Presidential', data3[, -1])
data4 <- cbind(data4[, 1, drop = FALSE], 'Type of Election' = 'House', data4[, -1])

# Remove type of election from all cells in the title row
colnames(data1) <- gsub("Pres Candidate ", "", colnames(data1))
colnames(data2) <- gsub("House ", "", colnames(data2))
colnames(data3) <- gsub("Pres Candidate ", "", colnames(data3))
colnames(data4) <- gsub("House ", "", colnames(data4))

# Add a "Year" column
data1 <- cbind(data1[, 1, drop = FALSE], Year = 2016, data1[, -1])
data2 <- cbind(data2[, 1, drop = FALSE], Year = 2018, data2[, -1])
data3 <- cbind(data3[, 1, drop = FALSE], Year = 2020, data3[, -1])
data4 <- cbind(data4[, 1, drop = FALSE], Year = 2022, data4[, -1])

# Remove year from all cells in the title row
colnames(data1) <- gsub("2016 ", "", colnames(data1))
colnames(data2) <- gsub("2018 ", "", colnames(data2))
colnames(data3) <- gsub("2020 ", "", colnames(data3))
colnames(data4) <- gsub("2022 ", "", colnames(data4))

# Merge cleaned dataframes
clean_age_data <- rbind(data1, data2, data3, data4)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
# VISUALIZATION 3 CLEANING

education_data <- subsets[["Education"]]
education_data <- education_data %>%
  mutate(`Type of Voter` = ifelse(grepl("Education", `Type of Voter`), NA, `Type of Voter`)) %>%
  filter_all(any_vars(!is.na(.)))

# Split the dataframe by election/year for individual cleaning
data1 <- education_data[, c("Type of Voter", "2016 D Pres Candidate Vote", "2016 D Pres Candidate Voter Share", 
                  "2016 D-R Pres Candidate Margin", "2016 Total Nonvoter Share", 
                  "2016 R Pres Candidate Vote", "2016 R Pres Candidate Voter Share", 
                  "2016 Total Voter Share")]
data2 <- education_data[, c("Type of Voter", "2018 D House Vote", "2018 D Voter Share", 
                  "2018 D-R House Margin", "2018 Total Nonvoter Share", 
                  "2018 R House Vote", "2018 R Voter Share", 
                  "2018 Total Voter Share")]

data3 <- education_data[, c("Type of Voter", "2020 D Pres Candidate Vote", "2020 D Pres Candidate Voter Share", 
                  "2020 D-R Pres Candidate Margin", "2020 Total Nonvoter Share", 
                  "2020 R Pres Candidate Vote", "2020 R Pres Candidate Voter Share", 
                  "2020 Total Voter Share")]

data4 <- education_data[, c("Type of Voter", "2022 D House Vote", "2022 D Voter Share", 
                  "2022 D-R House Margin", "2022 Total Nonvoter Share", 
                  "2022 R House Vote", "2022 R Voter Share", 
                  "2022 Total Voter Share")]

# Add a "Type of Election" column
data1 <- cbind(data1[, 1, drop = FALSE], 'Type of Election' = 'Presidential', data1[, -1])
data2 <- cbind(data2[, 1, drop = FALSE], 'Type of Election' = 'House', data2[, -1])
data3 <- cbind(data3[, 1, drop = FALSE], 'Type of Election' = 'Presidential', data3[, -1])
data4 <- cbind(data4[, 1, drop = FALSE], 'Type of Election' = 'House', data4[, -1])

# Remove type of election from all cells in the title row
colnames(data1) <- gsub("Pres Candidate ", "", colnames(data1))
colnames(data2) <- gsub("House ", "", colnames(data2))
colnames(data3) <- gsub("Pres Candidate ", "", colnames(data3))
colnames(data4) <- gsub("House ", "", colnames(data4))

# Add a "Year" column
data1 <- cbind(data1[, 1, drop = FALSE], Year = 2016, data1[, -1])
data2 <- cbind(data2[, 1, drop = FALSE], Year = 2018, data2[, -1])
data3 <- cbind(data3[, 1, drop = FALSE], Year = 2020, data3[, -1])
data4 <- cbind(data4[, 1, drop = FALSE], Year = 2022, data4[, -1])

# Remove year from all cells in the title row
colnames(data1) <- gsub("2016 ", "", colnames(data1))
colnames(data2) <- gsub("2018 ", "", colnames(data2))
colnames(data3) <- gsub("2020 ", "", colnames(data3))
colnames(data4) <- gsub("2022 ", "", colnames(data4))

# Merge cleaned dataframes
clean_education_data <- rbind(data1, data2, data3, data4)
```

# TRENDS IN VOTER TURNOUT
As shown in the chart below, voter turnout disparities persist across racial, educational, and gender lines. While there is some variation from election to election, these trends have remained remarkably consistent over the past four election cycles. This indicates that voter turnout is *not* random but rather influenced by structural and demographic factors.

```{r fig.width=10, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE}
# Define the desired order of the legend items
desired_order <- c("Total", "18-49", "50+", "Black, non-Hispanic", "White, non-Hispanic", 
                   "Hispanic", "some college or less", "college grad+", "not married", "married")

# Define your color scheme (unchanged)
colors <- c("Total" = "#051224",
            "18-49" = "#C6342F",
            "50+" = "#F84947",
            "Black, non-Hispanic" = "#004AAD",
            "White, non-Hispanic" = "#4689D0",
            "Hispanic" = "#B1DFFC",
            "some college or less" = "#808080",
            "college grad+" = "#ADADAD",
            "not married" = "#7C4A7A",
            "married" = "#BA8CB9")

# Common Theme for Both Plots
custom_theme <- theme_minimal() +
  theme(
    # Title and subtitle
    plot.title = element_text(hjust = 0.5, family = "bodoni_moda", size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, family = "montserrat", size = 12),
    
    # Axis labels and text
    axis.title = element_text(family = "montserrat", size = 14),
    axis.text = element_text(family = "montserrat", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    
    # Background styling (white background to match previous graph)
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    
    # Strip text for facets
    strip.text = element_text(family = "montserrat", size = 12, face = "bold"),
    
    # Legend customization
    legend.title = element_text(family = "bodoni_moda", size = 13, face = "bold"),
    legend.text = element_text(family = "montserrat", size = 12),
    legend.position = 'right',
    legend.title.align = 0.5,
    
    # Margins
    plot.margin = margin(20, 20, 20, 20)
  )

# Female Voter Population Plot
female_plot <- ggplot(female_voters, aes(x = Year, y = `Total Nonvoter Share`, 
                                         color = factor(`Type of Voter`, levels = desired_order))) +
  geom_line(size = 1.2) +  # Match previous graph line thickness
  geom_point(size = 2) +   # Add points for clarity
  labs(x = "Year", y = "Share of Nonvoters (%)", color = "Type of Voter") +
  scale_color_manual(values = colors) +
  custom_theme +
  ggtitle("Female Voter Population")

# Male Voter Population Plot
male_plot <- ggplot(male_voters, aes(x = Year, y = `Total Nonvoter Share`, 
                                     color = factor(`Type of Voter`, levels = desired_order))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(x = "Year", y = "Share of Nonvoters (%)") +
  scale_color_manual(values = colors) +
  custom_theme +
  theme(legend.position = "none") +  # Hide legend for male plot
  ggtitle("Male Voter Population")

# Combine Plots with Patchwork
patch <- male_plot + female_plot + 
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Nonvoter Trends Remain Consistent Across Recent Elections",
    theme = theme(plot.title = element_text(family = "bodoni_moda", size = 20, face = "bold", hjust = 0.5))
  )

# Display the patchwork
patch
```

Recognizing these demographic trends, the Democratic campaign has a strategic opportunity to strengthen its voter base by targeting and mobilizing nonvoters. The data highlights that young and less-educated individuals consistently make up the largest share of nonvoters, particularly among women. Since these patterns have remained stable over multiple election cycles, campaigns must address the systemic barriers that prevent these groups from voting, such as lack of access to resources, political disengagement, and economic instability.

As identified in an article by [PBS News](https://www.pbs.org/newshour/politics/voter-turnout-united-states?utm_source=chatgpt.com), several factors contribute to low voter turnout, including restrictive voting laws, lack of competitive elections, voter disillusionment, and logistical challenges. No single cause fully explaining the trend. However, to address these challenges and close the participation gap, campaigns can implement voter education efforts, grassroots organizing, and digital outreach strategies tailored to the demographics most heavily impacted. Addressing these obstacles could significantly increase turnout without necessarily needing to convert opposition voters.

If the Democratic campaign successfully engages nonvoters in 2024, it could expand its electoral advantage by broadening its base. Understanding what motivates these groups to vote—whether through policy concerns, community-driven activism, or outreach from trusted sources—will be critical. Specifically, this chart identifies young and low-educated individuals as the two largest shares of nonvoters, especially within the female population. How do these groups vote when they do turn out?

# TRENDS IN LOW-EDUCATED VOTER PREFERENCES
Voters with some college education or less have consistently represented the largest subgroup of nonvoters over the past four election cycles. However, voter turnout among this group increased in 2022 increased, suggesting that engagement efforts may be beginning to have an effect. These demographic shifts point to a key opportunity for campaigns to target and mobilize nonvoters, especially those with lower levels of education, as a way to expand their base and increase turnout.

```{r fig.width=10, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE}
# Filtering the data to include only voters with "Some college" or "HS or Less" education
educ_subset <- clean_education_data %>%                 
  filter(`Type of Voter` %in% c("HS or less", "Some college")) %>%
  # Calculating the percentage of "Other" votes
  mutate(`Other Vote` = 100 - `D Vote` - `R Vote`) %>%
  # Reshaping the data into longer format for plotting
  pivot_longer(cols = c("D Vote", "R Vote", "Other Vote"),   
               names_to = "Party", values_to = "Votes") %>%
  # Reordering the levels of the 'Party' factor variable
  mutate(Party = factor(Party, levels = c("Other Vote", "D Vote", "R Vote")))

# Plot
ggplot(educ_subset, aes(x = Year, y = Votes, fill = Party)) +
  geom_bar(stat = "identity", position = "stack") +
  
  # Labels inside bars (except "Other Vote" category)
  geom_text(data = subset(educ_subset, Party != "Other Vote"), 
            aes(label = paste0(round(Votes), "%")),  
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white", fontface = "bold", 
            family = "montserrat") + 

  # Facet wrap with custom labels
  facet_wrap(~`Type of Voter`, scales = "free_y", 
             labeller = labeller(`Type of Voter` = c("HS or less" = "High School or Less", 
                                                     "Some college" = "Some College"))) +

  # Titles and Labels
  labs(title = "Republican Gains Among Low-Educated Voters Stay Competitive",
       x = "Year", y = "Share of Votes (%)", fill = "Party") +

  # X-axis breaks for every 2 years
  scale_x_continuous(breaks = seq(min(educ_subset$Year), max(educ_subset$Year), by = 2)) +

  # Color scheme (unchanged)
  scale_fill_manual(values = c("#ADADAD", "#004AAD", "#F84947"), 
                    labels = c("Other", "Democrat", "Republican")) +

  # Theme adjustments for consistency
  theme_minimal() +
  theme(
    # Title and subtitle styling
    plot.title = element_text(family = "bodoni_moda", size = 16, face = "bold", hjust = 0.5), 
    
    # Axis titles and text
    axis.title = element_text(family = "montserrat", size = 14, face = "italic"),
    axis.text = element_text(family = "montserrat", size = 10),
    
    # Panel background (set to white for consistency)
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),

    # Facet strip text
    strip.text = element_text(family = "montserrat", size = 12, face = "bold"),

    # Legend adjustments
    legend.title = element_text(family = "bodoni_moda", size = 13, face = "bold"),
    legend.text = element_text(family = "montserrat", size = 12),
    legend.position = "right"  # Legend on right, no box
  )

```

The stacked bar chart above illustrates voting patterns among individuals with a high school education or less and those with some college education. The data shows that while these voters, particularly those with a high school education or less, predominantly support the Republican Party, the margin remains competitive across each election cycle. Although the Republican Party has gained ground among low-educated voters, the trend is not overwhelmingly one-sided. The Democratic Party still holds a substantial share of support, though turnout from this group remains an issue.

Given that the Democratic Party traditionally receives more support from low-educated voters, especially those with high school education or less, the lack of turnout among these voters could explain the Republican Party's increasing dominance in this group. To address this, campaigns should focus on engaging these voters by targeting the root causes of political disengagement, such as economic instability, limited access to resources, and political disillusionment. By enhancing voter education efforts and increasing outreach to this demographic, the Democrats could shift the trend in their favor in future elections.

# TRENDS IN YOUNG VOTER PREFERENCES
Young voters, the second largest subgroup of nonvoters, are a key demographic that overwhelmingly supports the Democratic Party. As the next Presidential Election will be a rematch of 2020, it’s insightful to examine the voting patterns of this age group in the previous election. The bar chart below displays the voter shares each candidate received from young voters aged 18-49 in 2020, offering a glimpse into the Democratic advantage among this group.

```{r fig.width=10, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE}
# Filtering the data for voters aged 18-49 in the 2020 election
age_subset <- clean_age_data %>%                 
  filter(`Type of Voter` %in% c("18-24", "25-29", "30-39", "40-49")) %>%
  filter(`Year` %in% c("2020")) %>%
  
  # Calculating the percentage of "Other" votes
  mutate(`Other Vote` = 100 - `D Vote` - `R Vote`) %>%
  
  # Convert "D Vote" values to negative numbers for visualization
  mutate(`D Vote` = -`D Vote`) %>%
  
  # Reshaping the data into a longer format for plotting
  pivot_longer(cols = c("D Vote", "R Vote", "Other Vote"),   
               names_to = "Party", values_to = "Votes") %>%
  
  # Reordering the levels of the 'Party' factor variable
  mutate(Party = factor(Party, levels = c("Other Vote", "D Vote", "R Vote"))) %>%
  
  # Add an "abs_votes" column with absolute values of "Votes"
  mutate(abs_votes = abs(Votes))

# Stacked horizontal bar chart
ggplot(age_subset, aes(x = factor(`Type of Voter`), y = Votes, fill = Party)) +
  
  # Bar chart
  geom_bar(stat = "identity") +
  
  # Labels inside bars (Only for D and R votes)
  geom_text(aes(label = ifelse(Party %in% c("D Vote", "R Vote"), paste0(abs_votes, "%"), "")),
            position = position_stack(vjust = 0.5), 
            color = "white", fontface = "bold", 
            family = "montserrat") + 
  
  # Titles and Labels
  labs(title = "Young Voters Favor the Democratic Party",
       x = "Age of Voter",
       y = "Share of Votes (%)",
       fill = "Presidential Candidate") + 
  
  # Theme adjustments for consistency
  theme_minimal() +
  theme(
    # Title styling
    plot.title = element_text(family = "bodoni_moda", size = 16, face = "bold", hjust = 0.5), 
    
    # Axis titles and text
    axis.title = element_text(family = "montserrat", size = 14, face = "italic"),
    axis.text = element_text(family = "montserrat", size = 10, color = "black"), # Ensuring black text for readability
    
    # Panel and background adjustments
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),

    # Legend adjustments (No box)
    legend.title = element_text(family = "bodoni_moda", size = 13, face = "bold"),
    legend.text = element_text(family = "montserrat", size = 12),
    legend.position = "right"  # Keep legend on the right
  ) +
  
  # Flip the coordinates for horizontal bars
  coord_flip() +

  # Custom fill colors (unchanged)
  scale_fill_manual(labels = c("Other", "Joe Biden", "Donald Trump"),
                    values = c("Other Vote" = "#ADADAD", "D Vote" = "#004AAD", "R Vote" = "#F84947"))
```

The chart highlights that, collectively, Democrats received majority support from all age groups between 18 and 49. However, the most significant support came from the youngest voters—those likely voting in their first Presidential election, aged 18-24. This group overwhelmingly backed Joe Biden, reflecting a strong preference for Democratic candidates.

To build on this momentum and continue securing the support of young voters, the Democratic Party can invest in targeted outreach programs, develop tailored messaging that resonates with their values, and further engage in community organizing initiatives. By maintaining and expanding this base of support, the Democrats can boost voter participation, grow their voter base, and potentially gain a decisive advantage come election day.

# CONCLUSION
Across the last four election cycles, patterns reveal that turnout is not random; rather, it correlates closely with identifiable demographic factors. ***With the 2024 General Election quickly approaching, the imperative for Democrats is clear: to secure victory, they must consolidate a formidable voter base characterized by robust turnout.*** This necessitates a strategic approach that identifies and engages eligible voters within their base who have historically exhibited lower levels of electoral participation, such as young voters. By deploying targeted campaign tactics aimed at motivating these segments of the electorate, the Democratic campaign can bolster its presence at the polls on Election Day. Additionally, by prioritizing outreach to previously marginalized constituents, such as low-educated voters, can help neutralize Republican advantages and potentially shift the balance in the Democrats' favor. Through these concerted efforts, Democrats can not only strengthen their electoral prospects but also uphold the foundational principles of representative democracy.